% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This TeX file is part of the course
%%%% Introduction to Scientific Programming in C++/Fortran2003
%%%% copyright 2017-2022 Victor Eijkhout eijkhout@tacc.utexas.edu
%%%%
%%%% address.tex : about addresses and references
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\Level 0 {Reference}
\label{sec:reference}

This section contains further facts about references, which you have
already seen as a mechanism for parameter passing;
section~\ref{sec:pass-by-ref}.
Make sure you study that material first.

Passing a variable to a routine passes the value; in the routine, the
variable is local.

\verbatimsnippet{localparm}

If you do want to make the change visible in the
%
\indexterm{calling environment}, use a reference:
%
\verbatimsnippet{refparm}
%
There is no change to the calling program. (Some people who are used
to~C find this bad,
since you can not see from the use of a function whether it passes
%
\emph{by reference}\index{parameter!passing!by reference}
%
or \emph{by value}\index{parameter!passing!by value}.)

\Level 0 {Pass by reference}

If you use a mathematical style of subprograms, where some values go
in, and a new entity comes out, in effect all the inputs can be
copied. This style is called
\indexterm{functional programming}, and
there is much to be said for it. For instance, it makes it possible
for the compiler to reason about your program.
The only thing you have to worry
about is the cost of copying, if the inputs are of non-trivial size,
such as arrays.

However, sometimes you want to alter the inputs, so instead of a copy
you need a way of accessing the actual input object. That's what
\emph{references}\index{reference} are invented for: to allow a subprogram access to the
actual input entity.

A bonus of using references is that you do not incur the cost of
copying. So what if you want this efficiency, but your program is
really functional in design? Then you can use a
\indextermsub{const}{reference}: the argument is passed by reference,
but you indicate explicitly that the subprogram does not alter it,
again allowing compiler optimizations.

\begin{block}{Reference: change argument}
\label{sl:refarg-change}
A reference makes the function parameter a synonym of the argument.
\begin{lstlisting}
void f( int &i ) { i += 1; };
int main() {
  int i = 2;
  f(i); // makes it 3
\end{lstlisting}
\end{block}

\begin{block}{Reference: save on copying}
\label{sl:refarg-nocopy}
\begin{multicols}{2}
\begin{lstlisting}
class BigDude {
public:
   vector<double> array(5000000);
}

void f(BigDude d) {
  cout << d.array[0];
};

int main() {
   BigDude big;
   f(big); // whole thing is copied
\end{lstlisting}
\columnbreak
Instead write:
\begin{lstlisting}
void f( BigDude &thing ) { .... };
\end{lstlisting}
Prevent changes:
\begin{lstlisting}
void f( const BigDude &thing ) { .... };
\end{lstlisting}
\hbox{}\vfill\hbox{}
\end{multicols}
\end{block}

\Level 0 {Reference to class members}
\label{sec:class-ref}

Here is the naive way of returning a class member:
\begin{lstlisting}
class Object {
private:
  SomeType thing;
public:
  SomeType get_thing() {
    return thing; };
};
\end{lstlisting}
The problem here is that the return statement makes a copy of
\n{thing}, which can be expensive. Instead, it is better to return the
member by \emph{reference}\index{reference!to class member}:
\begin{lstlisting}
  SomeType &get_thing() { 
    return thing; };
\end{lstlisting}
The problem with this solution is that the calling program can now
alter the private member. To prevent that, use a
\emph{const reference}\index{reference!const!to class member}:
%
\snippetwithoutput{constref}{const}{constref}

In the above example, the function giving a reference was used in the left-hand side of
an assignment. If you would use it on the right-hand side, you would
not get a reference. The result of an expression can not be a
reference.

Let's again make a class where we can get a reference to the
internals:
%
\verbatimsnippet{rhsrefclass}

Now we explore various ways of using that reference on the right-hand
side:
%
\snippetwithoutput{rhsref}{func}{rhsref}
%
(On the other hand, after \n{const auto &ref} the reference is not
modifiable. This variant is useful if you want read-only access,
without the cost of copying.)

You see that, despite the fact that the method \n{data} was defined as
returning a reference, you still need to indicate whether the
left-hand side is a reference.

See section~\ref{sec:constparam} for the interaction between \n{const}
and references.

\Level 0 {Reference to array members}
\label{sec:overloadbracket}

You can define various operator, such as \verb.+-*/. arithmetic
operators, to act on classes, with your own provided implementation;
see section~\ref{sec:operatordef}. You can also define the parentheses
and square brackets operators, so make your object look like a
function or an array respectively.

These mechanisms can also be used to provide safe access to arrays
and/or vectors that are private to the object.

Suppose you have an object that contains an \n{int} array. You can
return an element by defining the subscript (square bracket) operator
for the class:
%
\verbatimsnippet{refindexcopy}
%
Note that \n{return array[i]} will return a copy of the array element,
so it is not possible to write
\begin{lstlisting}
myobject[5] = 6;
\end{lstlisting}
For this we need to return a reference to the array element:
%
\verbatimsnippet{refindexref}

Your reason for wanting to return a reference could be to prevent the
\emph{copy of the return result}\index{return!makes copy}
that is induced by the \n{return} statement.
In this case, you may not want to be able to alter the object
contents, so you can return a \indextermsub{const}{reference}:
%
\verbatimsnippet{refindexconstref}

\Level 0 {rvalue references}

See the chapter about obscure stuff; section~\ref{sec:rvalue-ref}.
