% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This TeX file is part of the course
%%%% Introduction to Scientific Programming in C++/Fortran2003
%%%% copyright 2017-2022 Victor Eijkhout eijkhout@tacc.utexas.edu
%%%%
%%%% if.tex : conditionals
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A program consisting of just a list of assignment and expressions
would not be terribly versatile. At least you want to be able to say
`if some condition, do one computation, otherwise compute something else',
or: `until some test is true, iterate the following computations'.
The mechanism for
testing and choosing an action accordingly is called a
\indexterm{conditional}.
(Iterating is discussed in chapter~\ref{ch:loop}.)

\Level 0 {Conditionals}
\label{sec:if}

Here are some forms a conditional can take.

A single statement, executed if the test is true:
\begin{lstlisting}
if (x<0)
  x = -x;
\end{lstlisting}
Single statement in the true branch, and likewise a single
statement in the false branch:
\begin{lstlisting}
if (x>=0)
  x = 1;
else
  x = -1;
\end{lstlisting}
Both in the true and the false branch multiple statements are allowed,
if you enclose them in curly braces:
\begin{lstlisting}
if (x<0) {
  x = 2*x; y = y/2;
} else {
  x = 3*x; y = y/3;
}
\end{lstlisting}
You can chain conditionals by extending the \lstinline{else} part.
In this example the dots stand for omitted code:
\begin{lstlisting}
if (x>0) {
  ....
} else if (x<0) {
  ....
} else {
  ....
}
\end{lstlisting}
Conditionals can also be nested:
\begin{lstlisting}
if (x>0) {
  if (y>0) {
    ....
  } else {
    ....
  }
} else {
  ....
}
\end{lstlisting}
\begin{itemize}
\item
  In that last example the outer curly brackets in the true branch are
  optional. But it's safer to use them anyway.
\item When you start nesting constructs, use indentation to make it
  clear which line is at which level. A~good editor helps you with that.
\end{itemize}

\begin{slide}{If-then-else}
  \label{sl:ifthenelse}
  A \indextermdef{conditional} is a test: `if something is true, then do
  this, otherwise maybe do something else'. The C++ syntax is
\begin{lstlisting}
if ( something ) {
  // do something;
} else {
  // do otherwise;
}
\end{lstlisting}
\begin{itemize}
\item The `else' part is optional
\item You can leave out braces in case of single statement.
\end{itemize}
\end{slide}

\begin{slide}{Complicated conditionals}
  \label{sl:elseif}
  Chain:
\begin{lstlisting}
if ( /* some test */ ) {
  ...
} else if ( /* other test */ ) {
  ...
}
\end{lstlisting}
Nest:
\begin{lstlisting}
if ( /* some test */ ) {
  if ( /* other test */ ) {
    ...
  } else {
    ...
  }
}
\end{lstlisting}
\end{slide}

\begin{exercise}
  \label{ex:if-scope-ab}
  For what values of $x$ will the left code print~`\n{b}'?\\
  For what values of $x$ will the right code print~`\n{b}'?
  \begin{multicols}{2}
\begin{lstlisting}
float x = /* something */
if ( x > 1 ) {
   cout << "a" << endl;
   if ( x > 2 )
    cout << "b" << endl;
}
\end{lstlisting}
\columnbreak
\begin{lstlisting}
float x = /* something */
if ( x > 1 ) {
   cout << "a" << endl;
} else if ( x > 2 ) {
    cout << "b" << endl;
}
\end{lstlisting}
  \end{multicols}
\end{exercise}

\Level 0 {Operators}

You have already seen arithmetic expressions; now we need to look at
logical expressions: just what can be tested in a conditional. 
For the most part, logical expressions are intuitive.
However, note that they can be chained only in certain ways:
\begin{lstlisting}
bool x,y,z;
if ( x or y or z ) ; //good
int i,j,k;
if ( i < j < k ) ; // WRONG
\end{lstlisting}

\begin{block}{Comparison and logical operators}
  \label{sl:operators}
  Here are the most common \indextermsub{logic}{operator}s and
  \indextermsub{comparison}{operator}s:
  
  \begin{tabular}{lll}
    \toprule
    Operator&meaning&example\\ \midrule
    \n{==}&equals&\lstinline^x==y-1^\\
    \n{!=}&not equals&\lstinline^x*x!=5^\\
    \n{>}&greater&\lstinline^y>x-1^\\
    \n{>=}&greater or equal&\lstinline^sqrt(y)>=7^\\
    \n{<},\n{<=}&less, less equal&\\
    \n{&&},\n{||}&and, or&\lstinline^x<1 && x>0^\\
    \n{and},\n{or}&and, or&\lstinline^x<1 and x>0^\\
    \n{!}&not&\lstinline^!( x>1 && x<2 )^\\
    \n{not}&&\lstinline^not ( x>1 and x<2 )^\\
    \bottomrule
  \end{tabular}

  \emph{Precedence}\index{operator!precedence} rules of operators are
  common sense. When in doubt, use parentheses.
\end{block}

\begin{exercise}
  \label{ex:if2digits}
  The following code claims to detect if an integer has more than 2 digits.
  %
  \snippetwithoutput{if2digits}{basic}{if}
  %
  Fix the small error in this code.
  Also add an `else' part that prints if a number is negative.
  \skeleton{if}
\end{exercise}

\begin{exercise}
  \label{ex:oddeven}
  Read in an integer. If it is even, print `even', otherwise print
  'odd':
\begin{lstlisting}
if ( /* your test here */ ) 
  cout << "even" << '\n';
else
  cout << "odd" << '\n';
\end{lstlisting}
Then, rewrite your test so that the true branch corresponds to the odd case.
\end{exercise}

In exercise~\ref{ex:oddeven} it didn't matter whether you used the test for even
or for odd, but sometimes it does make a difference how you arrange complicated conditionals.
In the following exercise, think about how to arrange the tests. There is more than one way.

\begin{exercise}
  \label{ex:fizzbuzz}
  Read in a positive integer. If it's a multiple of three print `Fizz!';
  if it's a multiple of five print `Buzz!'.
  It is a multiple of both three and five print `Fizzbuzz!'.
  Otherwise print nothing.

  Note:
  \begin{itemize}
  \item Capitalization.
  \item Exclamation mark.
  \item Your program should display at most one line of output.
  \end{itemize}
\end{exercise}


\Level 1 {Bitwise logic}
\label{sec:bitc}

Above we only considered the \lstinline{and} and \lstinline{or} logical operators,
also spelled \n{&&} and \n{||}.
There are also \indextermsub{bitwise}{operator}s,
that look confusingly similar to the latter notation:

\snippetwithoutput{bitandor}{basic}{bitor}

To understand what's happening here, realize that
\[ \n{6}\equiv \n{110} \quad\hbox{and}\quad \n{3}\equiv\n{011} \]

You will probably not use the bitwise operators often, but
the following idiom is sometimes encountered:
\begin{lstlisting}
const int
  STATE_1 = 1, STATE_2 = 1<<1, STATE_3 = 1<<2;
int state = /* stuff */;
if ( state & ( STATE_1 | STATE_3 ) )
  cout << "We are in state 1 or 3";
\end{lstlisting}

\Level 1 {Review}

\begin{review}
  \label{q:if}
  True or false?
  \begin{itemize}
  \item The tests \lstinline$if (i>0)$ and \lstinline$if (0<i)$ are equivalent.
    \slackpollTF+Same tests: 'i>0' and '0<i' ?+
  \item The test
\begin{lstlisting}
if (i<0 && i>1) 
  cout << "foo"
\end{lstlisting}
    prints \n{foo} if
    $i<0$ and also if $i>1$.
    \slackpollTF+'if (i<0 && i>1)' is true if i negative and if i greater than one+
  \item The test 
\begin{lstlisting}
if (0<i<1)
  cout << "foo"
\end{lstlisting}
prints \n{foo} if $i$ is
    between zero and one.
    \slackpollTF+'if (0<i<1)' true if i between 0 and 1+
  \end{itemize}
\end{review}

\begin{review}
  \label{q:true}
  Any comments on the following?
\begin{lstlisting}
bool x;
// ... code with x ...
if ( x == true )
  // do something
\end{lstlisting}
  
\end{review}

\Level 0 {Switch statement}

If you have a number of cases corresponding to specific integer
values, there is the \indexc{switch} statement.

\begin{block}{Switch statement example}
  \label{sl:switch}
  Cases are executed consecutively until you `break' out of the switch
  statement:
  %
  \renewcommand\snippetcodefraction{.5}
  \renewcommand\snippetanswfraction{.48}
  \snippetwithoutput{switchstatement}{basic}{switch}
\end{block}

\begin{exercise}
  \label{ex:switch-range}
  Suppose the variable \n{n} is a nonnegative integer. Write a
  \n{switch} statement that has the same effect as:
\begin{lstlisting}
if (n<5)
  cout << "Small" << endl;
else
  cout << "Not small" << endl;
\end{lstlisting}
\end{exercise}

It is possible that the compiler generates more efficient code
from a \indexc{switch} statement than from a conditional.
Otherwise, there are no things you can do with a \indexc{switch}
that you can not do with a conditional.

\Level 0 {Scopes}

The true and false branch of a conditional can consist of a single
statement, or of a block in curly brackets. Such a block creates a
%
\emph{scope}\index{scope!in conditional branches}
%
where you can define local variables.

\begin{lstlisting}
if ( something ) {
  int i; 
  .... do something with i
}
// the variable `i' has gone away.
\end{lstlisting}

See chapter~\ref{ch:scope} for more on scopes.

\begin{slide}{Local variables in conditionals}
  \label{sl:if-scope}
  The curly brackets in a conditional allow you to define local variables:
\begin{lstlisting}
if ( something ) {
  int i; 
  .... do something with i
}
// the variable `i' has gone away.
\end{lstlisting}
Good practice: only define variable where needed.

Braces induce a \indexterm{scope}.
\end{slide}

\Level 0 {Advanced topics}
\label{sec:if-advanced}

\Level 1 {Short-circuit evaluation}

C++ logic operators have a feature called
\indexterm{short-circuit evaluation}:
a logical operator stops evaluating
in strict left-to-right order
when the result is clear.
For instance, in
\begin{lstlisting}
clause1 and clause2
\end{lstlisting}
the second clause is not evaluated if the first one is
\lstinline{false}, because the truth value of this conjunction is
already determined.

Likewise, in
\begin{lstlisting}
clause1 or clause2
\end{lstlisting}
the second clause is not evaluated if the first one is \lstinline{true},
because the value of the \lstinline{or} conjunction is already clear.

This mechanism allows you to write
\begin{lstlisting}
if ( x>=0 and sqrt(x)<10 ) { /* ... */ }
\end{lstlisting}
Without short-circuit evaluation the square root operator could be applied to
negative numbers.

\Level 1 {Ternary if}

The true and false branch of a conditional contain whole statements.
For example
\begin{lstlisting}
if (foo)
  x = 5;
else
  y = 6;
\end{lstlisting}
But what about the case where the true and false branch
assign to the same variable, but with a different expression?
You can not write
\begin{multicols}{2}
Original code:
\begin{lstlisting}
if (foo)
  x = 5;
else
  x = 6;
\end{lstlisting}
\columnbreak
Not legal syntax for `simplification':
\begin{lstlisting}
x = if (foo) 5; else 6; 
\end{lstlisting}
\end{multicols}
For this case there is the \indextermsub{ternary}{if},
which acts as if it's an expression itself, but chosen
between two expressions.
The previous assignment to~\lstinline{x} then becomes:
\begin{lstlisting}
x = foo ? 5 : 6;
\end{lstlisting}
Surprisingly, this expression can even be in the left-hand side:
\begin{lstlisting}
foo ? x : y = 7;
\end{lstlisting}

\Level 1 {Initializer}
\label{sec:if-init}

The \cppstandard{17} standard introduced a new form of the \indexc{if} and
\indexc{switch} statement: it is possible to have a single
statement of declaration prior to the test.
This is called the
\emph{initializer}\index{initializer!in conditional}.

\snippetwithoutput{ifinitchar}{basic}{ifinit}

This is particularly elegant if the init statement is a declaration,
because the declared variable is then local to the conditional.
Previously one would have had to write
\begin{lstlisting}
char c;
c = getchar();
if ( c!='a' ) /* ... */
\end{lstlisting}
with the variable defined outside of the scope of the conditional.

\Level 0 {Review questions}

\begin{review}
  T/F: the following is a legal program:
\begin{lstlisting}
#include <iostream>
int main() {
  if (true)
    int i = 1;
  else
    int i = 2;
  std::cout << i;
  return 0;
}
\end{lstlisting}
\end{review}

\begin{review}
  T/F: the following are equivalent:
  \begin{multicols}{2}
\begin{lstlisting}
if (cond1)
  i = 1;
else if (cond2)
  i = 2;
else
  i = 3;
\end{lstlisting}
\columnbreak
compare:
\begin{lstlisting}
if (cond1)
  i = 1;
else {
  if (cond2)
    i = 2;
  else
    i = 3;
}
\end{lstlisting}
\end{multicols}
\end{review}
