% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This TeX file is part of the course
%%%% Introduction to Scientific Programming in C++/Fortran2003
%%%% copyright 2017-2020 Victor Eijkhout eijkhout@tacc.utexas.edu
%%%%
%%%% ifj.tex : conditionals in Julia
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A program consisting of just a list of assignment and expressions
would not be terribly versatile. At least you want to be able to say
`if $x>0$, do one computation, otherwise compute something else', or `until some
test is true, iterate the following computations'. The mechanism for
testing and choosing an action accordingly is called a
\indexterm{conditional}.

\Level 0 {Basic syntax}
\label{sec:ifj}

Conditionals are delimited by \lstinline{if} and \lstinline{end}.
\begin{lstlisting}
if x<0
  x = 2*x; y = y/2;
else
  x = 3*x; y = y/3;
end
\end{lstlisting}
Chaining conditionals (where the dots stand for omitted code):
\begin{lstlisting}
if x>0
  ....
elseif x<0
  ....
else
  ....
end
\end{lstlisting}
Nested conditionals:
\begin{lstlisting}
if x>0
  if y>0
    ....
  else
    ....
  end
else
  ....
end
\end{lstlisting}

When you start nesting constructs, use indentation to make it
  clear which line is at which level. A~good editor helps you with that.


\begin{exercise}
  \label{ex:if-scope-abj}
  For what values of $x$ will the left code print~`\n{b}'?\\
  For what values of $x$ will the right code print~`\n{b}'?
  \begin{multicols}{2}
\begin{lstlisting}
x = ## some float value
if x > 1
   println("a")
   if x > 2
     println("b")
   end
end
\end{lstlisting}
\columnbreak
\begin{lstlisting}
x = ## some float value
if x > 1
   println("a"
elseif x > 2 
   println("b")
end
\end{lstlisting}
  \end{multicols}
\end{exercise}

\begin{comment}
\Level 0 {Operators}

You have already seen arithmetic expressions; now we need to look at
logical expressions: just what can be tested in a conditional. 
Here is a fragment of language grammar that spells out what is
legal. You see that most of the rules are recursive, but there is an
important exception.

\begin{block}{What are logical expressions?}
  \label{sl:logical-syntax}
\begin{verbatim}
logical_expression :: 
  comparison_expression
  | NOT comparison_expression
  | logical_expression CONJUNCTION comparison_expression
comparison_expression :: 
  numerical_expression COMPARE numerical_expression
numerical_expression :: 
  quantity
  | numerical_expression OPERATOR quantity
quantity :: number | variable
\end{verbatim}
\end{block}

\begin{block}{Comparison and logical operators}
  \label{sl:operators}
  Here are the most common \indextermsub{logic}{operator}s and
  \indextermsub{comparison}{operator}s.
  \begin{tabular}{|l|l|l|}
    \hline
    Operator&meaning&example\\ \hline
    \texttt{==}&equals&\texttt{x==y-1}\\
    \texttt{!=}&not equals&\texttt{x*x!=5}\\
    \texttt{>}&greater&\texttt{y>x-1}\\
    \texttt{>=}&greater or equal&\texttt{sqrt(y)>=7}\\
    \texttt{<},\texttt{<=}&less, less equal&\texttt{}\\
    \n{&&},\n{||}&and, or&\n{x<1 && x>0}\\
    \n{and},\n{or}&&\n{x<1 and x>0}\\
    \texttt{!}&not&\n{!( x>1 && x<2 )}\\
    \texttt{not}&&\n{not ( x>1 and x<2 )}\\
    \hline
  \end{tabular}

  \emph{Precedence}\index{operator!precedence} rules of operators are
  common sense. When in doubt, use parentheses.
\end{block}

\begin{remark}
  There are also \indextermsub{bitwise}{operator}s.
\end{remark}

\begin{exercise}
  \label{ex:oddeven}
  Read in an integer. If it is even, print `even', otherwise print
  'odd':
\begin{lstlisting}
if ( /* your test here */ ) 
  cout << "even" << endl;
else
  cout << "odd" << endl;
\end{lstlisting}
Then, rewrite your test so that the true branch corresponds to the odd case?
\end{exercise}

\begin{exercise}
  \label{ex:fizzbuzz}
  Read in an integer. If it's a multiple of three print `Fizz!';
  if it's a multiple of five print `Buzz'!. It it is 
  a multiple of both three and five print `Fizzbuzz!'. Otherwise
  print nothing. (Note: your program should display at most one output.)
\end{exercise}

\begin{review}
  \label{q:if}
  True or false?
  \begin{itemize}
  \item The tests \lstinline$if (i>0)$ and \lstinline$if (0<i)$ are equivalent.
  \item The test
\begin{lstlisting}
if (i<0 && i>1) 
  cout << "foo"
\end{lstlisting}
prints \n{foo} if
    $i<0$ and also if $i>1$.
  \item The test 
\begin{lstlisting}
if (0<i<1)
  cout << "foo"
\end{lstlisting}
prints \n{foo} if $i$ is
    between zero and one.
  \end{itemize}
  Any comments on the following?
\begin{lstlisting}
bool x;
// ... code with x ...
if ( x == true )
  // do something
\end{lstlisting}
\end{review}

\Level 0 {Switch statement}

If you have a number of cases corresponding to specific integer
values, there is the \indextermtt{switch} statement.

\begin{block}{Switch statement example}
  \label{sl:switch}
  Cases are executed consecutively until you `break' out of the switch
  statement:
  %
  \snippetwithoutput{switchstatement}{basic}{switch}
\end{block}

\begin{exercise}
  \label{ex:switch-range}
  Suppose the variable \n{n} is a nonnegative integer. Write a
  \n{switch} statement that has the same effect as:
\begin{lstlisting}
if (n<5)
  cout << "Small" << endl;
else
  cout << "Not small" << endl;
\end{lstlisting}
\end{exercise}

\Level 0 {Scopes}

The true and false branch of a conditional can consist of a single
statement, or of a block in curly brackets. Such a block creates a
%
\emph{scope}\index{scope!in conditional branches}
%
where you can define local variables.

\begin{lstlisting}
if ( something ) {
  int i; 
  .... do something with i
}
// the variable `i' has gone away.
\end{lstlisting}

See chapter~\ref{ch:scope} for more detail.

\begin{slide}{Local variables in conditionals}
  \label{sl:if-scope}
  The curly brackets in a conditional allow you to define local variables:
\begin{lstlisting}
if ( something ) {
  int i; 
  .... do something with i
}
// the variable `i' has gone away.
\end{lstlisting}
Good practice: only define variable where needed.

Braces induce a \indexterm{scope}.
\end{slide}
\end{comment}

\Level 0 {Advanced topics}

\Level 1 {Short-circuit operators}

Julia logic operators have a feature called short-circuiting:
a \indextermsub{short-circuit}{operator} stops evaluating when the
result is clear. For instance, in
\begin{lstlisting}
clause1 and clause2
\end{lstlisting}
the second clause is not evaluated if the first one is
\lstinline{false}, because the truth value of this conjunction is
already determined.

Likewise, in
\begin{lstlisting}
clause1 or clause2
\end{lstlisting}
the second clause is not evaluated if the first one is \lstinline{true}.

This mechanism allows you to write
\begin{lstlisting}
if ( x>=0 and sqrt(x)<10 ) { /* ... */ }
\end{lstlisting}
Without short-circuiting the square root operator could be applied to
negative numbers.

\Level 1 {Ternary if}

The true and false branch of a conditional contain whole statements.
For example
\begin{lstlisting}
if (foo)
  x = 5;
else
  y = 6;
\end{lstlisting}
But what about the case where the true and false branch perform
in effect the same action, and only differ by an expression?
\begin{lstlisting}
if (foo)
  x = 5;
else
  x = 6;
\end{lstlisting}
For this case there is the \indextermsub{ternary}{if},
which acts as if it's an expression itself, but chosen
between two expressions.
The previous assignment to~\lstinline{x} then becomes:
\begin{lstlisting}
x = foo ? 5 : 6;
\end{lstlisting}
Surprisingly, this expression can even be in the left-hand side:
\begin{lstlisting}
foo ? x : y = 7;
\end{lstlisting}

\begin{comment}
\Level 1 {Init statements}

The \cppstandard{17} introduced a new form of the \indextermtt{if} and
\indextermtt{switch} statement: it is possible to have a single
statement of declaration prior to the test.

\snippetwithoutput{ifinitchar}{basic}{ifinit}

This is particularly elegant if the init statement is a declaration,
because the declared variable is then local to the
conditional. Previously one would have had to write
\begin{lstlisting}
char c;
c = getchar();
if ( c!='a' ) /* ... */
\end{lstlisting}
with the variable globally defined.
\end{comment}
