% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This TeX file is part of the course
%%%% Introduction to Scientific Programming in C++/Fortran2003
%%%% copyright 2017-2022 Victor Eijkhout eijkhout@tacc.utexas.edu
%%%%
%%%% inheritance.tex : about the relation between classes
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\Level 0 {Inclusion relations between classes}
\label{sec:hasa}

The data members of an object can be of elementary datatypes, or they
can be objects. For instance, if you write software to manage courses,
each \lstinline{Course} object will likely have a \lstinline{Person} object,
corresponding to the teacher.

\begin{lstlisting}
class Person {
  string name;
  ....
}
class Course {
private:
  int year;
  Person the_instructor;
  vector<Person> students;
}
\end{lstlisting}

Designing objects with relations between them
is a great mechanism for writing structured code,
as it makes the objects in code behave like objects in the real world.
The relation where one object contains another, is called a
\indexterm{has-a relation} between classes.

\begin{slide}{Has-a relationship}
  \label{sl:obj-hasa}
  A class usually contains data members. These can be simple types or
  other classes. This allows you to make structured code.
\begin{lstlisting}
class Person {
  string name;
  ....
};
class Course {
private:
  Person the_instructor;
  int year;
};`
\end{lstlisting}
This is called the \indexterm{has-a relation}:\\
\lstinline{Course} has-a \lstinline{Person}
\end{slide}

\Level 1 {Literal and figurative has-a}

Sometimes a class can behave as if it includes an object of another
class, while not storing this other object.
Consider the example of a line segment, that is, the segment
from a starting point to an ending point.
We want to offer the \ac{API}:
\begin{lstlisting}
int main() {
  Segment somesegment( /* something */ );
  Point somepoint = somesegment.get_the_end_point();
\end{lstlisting}

We can support this by letting the \lstinline{Segment} class
actually store the starting and ending points:
\begin{lstlisting}
class Segment {
private:
  Point starting_point,ending_point;
}
\end{lstlisting}
or letting it store a distance and angle from the starting point:
\begin{lstlisting}
class Segment {
private:
  Point starting_point;
  float length,angle;
}
\end{lstlisting}
In both cases the code using the object is written as if the segment
object contains two points.
This illustrates how object-oriented programming can decouple the
\ac{API} of classes from their actual implementation.

\begin{slide}{Literal and figurative has-a}
  \label{sl:obj-hasa-ish}
\small
  A line segment has a starting point and an end point.
\begin{multicols}{2}
  A \lstinline{Segment}
  class can store those points:
\begin{lstlisting}
class Segment {
private:
  Point starting_point,ending_point;
public:
  Point get_the_end_point() {
    return ending_point; };
}
int main() {
  Segment somesegment;
  Point somepoint =
    somesegment.get_the_end_point();
\end{lstlisting}
\columnbreak
or store one and derive the other:
\begin{lstlisting}
class Segment {
private:
  Point starting_point;
  float length,angle;
public:
  Point get_the_end_point() {
    /* some computation
       from the
       starting point */ };
}
\end{lstlisting}
\vfill\hbox{}
\end{multicols}
Implementation vs API: implementation can be very different from user interface.
\end{slide}

Related to this decoupling, a class can also have two very different constructors.
\begin{block}{Multiple constructors}
  \label{sl:segment-constructors}
  \begin{lstlisting}
    class Segment {
      private:
      // up to you how to implement!
      public:
      Segment( Point start,float length,float angle )
      { .... }
      Segment( Point start,Point end ) { ... }
  \end{lstlisting}
  Depending on how you actually implement the class, the one constructor
  will simply store the defining data, and the other will do some
  conversion from the given data to the actually stored data.
\end{block}

This is another strength of object-oriented programming: you can
change your mind about the implementation of a class without having
to change the program that uses the class.

When you have a has-a relation between classes, the `default
constructor' problem (section~\ref{sec:default-constructor}) can pop
up again:

\begin{block}{Constructors in has-a case}
  \label{sl:hasa-construct-1}
\begin{multicols}{2}
Class for a person:
\begin{lstlisting}
class Person {
private:
  string name;
public:
  Person( string name ) {
    /* ... */
  };
};
\end{lstlisting}
\columnbreak
Class for a course, which contains a person:
\begin{lstlisting}
class Course {
private:
  Person instructor;
  int enrollment;
public:
  Course( string instr,int n ) {
    /* ???? */
  };
};
\end{lstlisting}
\end{multicols}
You want to use this as \lstinline+Course("Eijkhout",65);+
\end{block}

\begin{block}{Constructors in the has-a case}
  \label{sl:hasa-construct-2}
  Possible constructor:
\begin{lstlisting}
Course( string teachername,int nstudents ) {
  instructor = Person(teachername);
  enrollment = nstudents;
};
\end{lstlisting}
Preferred:
\begin{lstlisting}
Course( string teachername,int nstudents )
  : instructor(Person(teachername)),
    enrollment(nstudents) {
};
\end{lstlisting}
\end{block}

\begin{block}{Default Constructors in initialization}
  \label{sl:member-init-def-construct}
\begin{lstlisting}
class Inner { /* ... */ };
class Outer {
private:
  Inner inside_thing;
\end{lstlisting}
Two possibilities for constructor:
\begin{multicols}{2}
\begin{lstlisting}
Outer( Inner thing )
: inside_thing(thing) {};
\end{lstlisting}
The \lstinline{Inner} object is copied during construction
of \lstinline{Outer} object.
\par\vfill\hbox{}

\columnbreak

\begin{lstlisting}
Outer( Inner thing ) {
  inside_thing = thing;
};
\end{lstlisting}
The \lstinline{Outer} object is created, including construction of
\lstinline{Inner} object, then the argument is copied into place:
$\Rightarrow$~needs default constructor on \lstinline{Inner}.

\end{multicols}
\end{block}

\begin{slide}{Polymorphism in constructors}
  \label{sl:obj-poly-construct}
  You have to decide what to store and what to derive, but you can
  construct two ways:
\begin{lstlisting}
class Segment {
private:
 // up to you how to implement!
public:
  Segment( Point start,float length,float angle )
    { .... }
  Segment( Point start,Point end ) { ... }
\end{lstlisting}
Advantage: with a good API you can change your mind about the
implementation without changing the calling code.
\end{slide}

\begin{exercise}
  If you are doing the geometry project, this is a good time to
  do the exercises in section~\ref{sec:poly-rectangle}.
\end{exercise}

\Level 2 {Shorthand for objects}

For classes with a constructor, you can use a shorthand
for an object, giving a brace-delimited initializer list.

\begin{block}{Initializer lists}
  \label{sl:class-inlist}
  \emph{Initializer lists}\index{initializer list} can be used as denotations.
  \verbatimsnippet{initcurlypoint}  
\end{block}

\Level 0 {Inheritance}
\label{sec:inheritance}

In addition to the has-a relation, there is the \indexterm{is-a
  relation}, also called \indextermdef{inheritance}. Here one class is
a special case of another class.
Typically the object of the \indextermsub{derived}{class} (the special
case) then also inherits the data and methods of the
\indextermsub{base}{class} (the general case).

\begin{block}{Examples for base and derived cases}
  \label{sl:derived-example}
  \begin{itemize}
  \item Base case: employee. Has: salary, employee number.\\
    Special case: manager. Has in addition: underlings.
  \item Base case: shape in drawing program. Has: extent, area,
    drawing routine.\\
    Special case: square et cetera; has specific drawing routine.
  \end{itemize}  
\end{block}

How do you define a derived class?
The general code schema for use of a base class and derived class goes like this:

\begin{multicols}{2}
Base class, general case:
\begin{lstlisting}
class General {
protected: // note!
 int g;
public:
 void general_method() {};
};
\end{lstlisting}
\columnbreak
Derived class, special case:
\begin{lstlisting}
class Special : public General {
public:
  void special_method() { ... g ... };
};
\end{lstlisting}
\end{multicols}

\begin{slide}{General case, special case}
  \label{sl:obj-case}
  You can have classes where an object of one class is a special case of
  the other class. You declare that as
\begin{lstlisting}
class General {
protected: // note!
 int g;
public:
 void general_method() {};
};

class Special : public General {
public:
  void special_method() { g = ... };
};
\end{lstlisting}
\end{slide}

These are the various aspects of declaring a derived class:
\begin{itemize}
\item You need to indicate what the base class is:
\begin{lstlisting}
class Special : public General { .... }
\end{lstlisting}
\item The base class needs to declare its data members as
  \indexcdef{protected}: this is similar to private, except that
  they are visible to derived classes
\item The methods of the derived class can then refer to data of the
  base class;
\item Any method or data member defined for the base class is available
  for a derived class object.
\end{itemize}

\begin{slide}{Inheritance: derived classes}
  \label{sl:obj-derive}
  \emph{Derived}\index{class!derived} class \lstinline{Special}
  \emph{inherits}\index{inheritance} methods and data from
  \indextermsub{base}{class} \lstinline{General}:
\begin{lstlisting}
int main() {
  Special special_object;
  special_object.general_method();
  special_object.special_method();
}
\end{lstlisting}
Members of the base class need to be \lstinline{protected},
not \lstinline{private}, to be inheritable.
\end{slide}

The derived class has its own constructor, with the same name as the
class name, but when it is invoked, it also calls the constructor of
the base class. This can be the default constructor, but often you
want to call the base constructor explicitly, with parameters that are
describing how the special case relates to the base case.

In the following example, we have a general case, depending on two
independent parameters.
The special case comes from having a certain relationship between
these parameters.

\begin{lstlisting}
class General {
public:
  General( double x,double y ) {};
};
class Special : public General {
public:
  Special( double x ) : General(x,x+1) {};
};
\end{lstlisting}

\begin{slide}{Constructors}
  \label{sl:obj-derive-construct}
  When you run the special case constructor, usually the general
  constructor needs to run too. By default the `default constructor',
  but usually explicitly invoked:
\begin{lstlisting}
class General {
public:
  General( double x,double y ) {};
};
class Special : public General {
public:
  Special( double x ) : General(x,x+1) {};
};
\end{lstlisting}
\end{slide}

\begin{block}{Access levels}
  \label{sl:private-etc}
  Methods and data can be
  \begin{itemize}
  \item \lstinline{private}, because they are only used internally;
  \item \lstinline{public}, because they should be usable from outside a class
    object, for instance in the main program;
  \item \lstinline{protected}, because they should be usable in derived classes.
  \end{itemize}
\end{block}

\begin{exercise}
  If you are doing the geometry project, 
  you can now do the exercises in section~\ref{sec:geom-isa}.
\end{exercise}

\Level 1 {Methods of base and derived classes}
\label{sec:derive-method}

Above, it was assumed that derived classes use the methods
of the base class unchanged.
Sometimes, however, you may want the derived class have a different version
of a method.
This is done through the \indexc{virtual} and \indexc{override} keywords.

\begin{block}{Overriding methods}
  \label{sl:obj-method-override}
  \begin{itemize}
  \item A derived class can inherit a method from the base class.
  \item A derived class can define a method that the base class does
    not have.
  \item A derived class can \emph{override}\index{method!overriding} a
    base class method:
\begin{lstlisting}
class Base {
public:
  virtual f() { ... };
};
class Deriv : public Base {
public:
  virtual f() override { ... };
};
\end{lstlisting}
  \end{itemize}
\end{block}

\begin{block}{Override and base method}
  \label{sl:virtualoverride}
  \snippetwithoutput{virtualoverride}{object}{virtual}
\end{block}

\Level 1 {Virtual methods}

The methods of base and derived classes can relate in a number of ways.

\begin{block}{Base vs derived methods}
  \begin{itemize}
  \item Method defined in base class: can be used in any derived class.
  \item Method define in derived class: can only be used in that
    particular derived class.
  \item Method defined both in base and derived class, marked
    \indexc{override}: derived class method replaces (or extends)
    base class method.
  \item Virtual method: base class only declares that a routine has to
    exist, but does not give base implementation.\\ A class is called
    \indextermsub{abstract}{class} if it has virtual methods; pure
    virtual if all methods are virtual.\\ You can not make abstract objects.
  \end{itemize}  
\end{block}

\begin{block}{Abstract classes}
  \label{sl:abstract-class}
  Special syntax for \indextermsub{abstract}{method}:
\begin{lstlisting}
class Base {
public:
  virtual void f() = 0;
};
class Deriv : public Base {
public:
  virtual void f() { ... };
};
\end{lstlisting}
\end{block}

\begin{block}{Example: using virtual class}
  \small
  \begin{multicols}{2}
    \verbatimsnippet{virtfuncvirt}
    \vfill\columnbreak
    Suppose \lstinline{DenseVector} derives from \lstinline{VirtualVector}:\\
    \verbatimsnippet{virtfuncmain}
  \end{multicols}
\end{block}

\begin{block}{Implementation}
  \footnotesize
  \begin{multicols}{2}
    \verbatimsnippet{virtfuncdense}
    %% \vfill\columnbreak
    %% \verbatimsnippet{virtfuncfunc}
  \end{multicols}
\end{block}

\begin{exercise}
  Advanced! Write a small `integrator' library for
  \acp{ODE}s. Assuming `autonomous \acp{ODE}', that is $u'=f(t)$ with
  no $u$-dependence, there are two simple integration schemes:
  \begin{itemize}
  \item explicit: $u_{n+1} = u_n+\Delta t\,f_n$; and
  \item implicit: $u_{n+1} = u_n+\Delta t\,f_{n+1}$.
  \end{itemize}
  Write an abstract \lstinline{Integrator} class where the
  \lstinline{nextstep} method (which integrates for another~$\Delta t$) is pure virtual;
  then write
  \lstinline{ExplicitIntegrator} and \lstinline{ImplicitIntegrator}
  classes deriving from this.

  \verbatimsnippet{integratefwd}

  You can hardcode the function to be integrated, or try to pass a function pointer.
\end{exercise}

\Level 1 {Friend classes}

\begin{block}{Friend classes}
  \label{sl:obj-friend}
  A \indextermdef{friend} class can access private data and methods
  even if there is no inheritance relationship.
\begin{lstlisting}
/* forward definition: */ class A;
class B {
  friend class A;
private:
  int i;
};
class  A {
public: 
  void f(B b) { b.i; };
};
\end{lstlisting}
\end{block}

\Level 1 {Multiple inheritance}

\begin{block}{More}
  \label{sl:obj-more}  
  \begin{itemize}
  \item  Multiple inheritance: an~X is-a~A, but also is-a~B.\\
    This mechanism is somewhat dangerous.
  \item Virtual base class: you don't actually define a function in
    the base class, you only say `any derived class has to define this
    function'.
  \end{itemize}
\end{block}

