% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This TeX file is part of the course
%%%% Introduction to Scientific Programming in C++/Fortran2003
%%%% copyright 2017-2022 Victor Eijkhout eijkhout@tacc.utexas.edu
%%%%
%%%% namespace.tex : about namespaces
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\Level 0 {Solving name conflicts}
\label{sec:usename}

In section~\ref{sec:stdvector} you saw that the C++ \ac{STL} comes with
a \indexterm{vector} class, that implements dynamic arrays. You say
\begin{lstlisting}
std::vector<int> bunch_of_ints;
\end{lstlisting}
and you have an object that can store a bunch of ints. And if you use
such vectors often, you can save yourself some typing by 
\indexc{using namespace}.
You put
\begin{lstlisting}
using namespace std;
\end{lstlisting}
somewhere high up in your file, and write
\begin{lstlisting}
vector<int> bunch_of_ints;
\end{lstlisting}
in the rest of the file.

More safe:
\begin{lstlisting}
using std::vector;
\end{lstlisting}

But what if you are writing a geometry package, which includes a
vector class? Is there confusion with the \ac{STL} vector class?
There would be if it weren't for the phenomenon
\indextermdef{namespace}, which acts as as disambiguating prefix for
classes, functions, variables.

\begin{slide}{You have already seen namespaces}
  \label{sl:vec-namespace}
  Safest:
\begin{lstlisting}
#include <vector>
int main() {
  std::vector<stuff> foo;
}
\end{lstlisting}
\begin{multicols}{2}
  Drastic:
\begin{lstlisting}
#include <vector>
using namespace std;
int main() {
  vector<stuff> foo;
}
\end{lstlisting}
\vfill\columnbreak
Prudent:
\begin{lstlisting}
#include <vector>
using std::vector;
int main() {
  vector<stuff> foo;
}
\end{lstlisting}
\end{multicols}
\end{slide}

You have already seen namespaces in action when you wrote
\n{std::vector}: the `\n{std}' is the name of the namespace.

\begin{block}{Defining a namespace}
  \label{sl:namespace-def}
  You can make your own namespace by writing
\begin{lstlisting}
namespace a_namespace {
  // definitions
  class an_object { 
  };
|
\end{lstlisting}
\end{block}

so that you can write
\begin{block}{Namespace usage}
  \label{sl:namespace-use}
Qualify type with namespace:
\begin{lstlisting}
a_namespace::an_object myobject();
\end{lstlisting}
or
\begin{lstlisting}
using namespace a_namespace;
an_object myobject();
\end{lstlisting}
or
\begin{lstlisting}
using a_namespace::an_object;
an_object myobject();
\end{lstlisting}
or
\begin{lstlisting}
using namespace abc = space_a::space_b::space_c;
abc::func(x)
\end{lstlisting}
\end{block}

\Level 1 {Namespace header files}

If your namespace is going to be used in more than one program, you
want to have it in a separate source file, with an accompanying header
file:
%
\begin{block}{Including and using a namespace}
  \label{sl:nameinclude}
  There is a \lstinline{vector} in the standard namespace
  and in the new \lstinline{geometry} namespace:
  %
  \verbatimsnippet{nameinclude}
\end{block}

The header would contain the normal function and class headers, but
now inside a named namespace:
%
\begin{block}{Header definition}
  \label{sl:nameheader}
  \verbatimsnippet{nameheader}
\end{block}

and the implementation file would have the implementations,
in a
namespace of the same name:
%
\begin{block}{Implementations}
  \label{sl:nameimpl}
  \verbatimsnippet{nameimpl}
\end{block}

\Level 0 {Namespaces and libraries}

As the introduction to this chapter argued,
namespaces are a good way to prevent name conflicts.
This means that it's a good idea to create a namespace
for all your routines.
You see this design in almost all published C++ libraries.

Now consider this scenario:
\begin{enumerate}
\item You write a program that uses a certain library;
\item A new version of the library is released and installed on your system;
\item Your program, using shared/dynamic libraries, starts using this library,
  maybe even without you realizing it.
\end{enumerate}
This means that the old and new libraries need to be compatible in several ways:
\begin{enumerate}
\item All the classes, functions, and data structures defined in the earlier version
  also need to be defined in the new. This is not a big problem: new library
  versions typically add functionality. However,
\item The data layout of the new version needs to be the same.
\end{enumerate}

That second point is subtle. To illustrate, consider the library
is upgraded:
\begin{multicols}{2}
  First version:
  \begin{lstlisting}
    namespace geometry {
      class vector {
        private:
        std::vector<float> data;
        std::string name;
      }
    }
  \end{lstlisting}
  \columnbreak
  New version:
  \begin{lstlisting}
    namespace geometry {
      class vector {
        private:
        std::vector<float> data;
        int id;
        std::string name;
      }
    }
  \end{lstlisting}
\end{multicols}

The problem is that your compiled program has explicit
information where the class members are located in the class object.
By changing that structure of the objects, those references are no longer correct.
This is called `\ac{ABI} breakage' and
it leads to \indexterm{undefined behavior}.

The library can prevent this by:
\begin{lstlisting}
  namespace geometry {
    inline namespace v1.0 {
      class vector {
        private:
        std::vector<float> data;
        std::string name;
      }
    }
  }
\end{lstlisting}
and updating the version number in future version.
The program using this library implicitly uses the namespace
\lstinline+geometry::v1.0::vector+
so after a library update,
trying to execute the program

\Level 0 {Best practices}
\label{sec:namespace-practice}

In this course we advocated
pulling in functions explicitly:
\begin{lstlisting}
#include <iostream>
using std::cout;
\end{lstlisting}

It is also possible to use
\begin{lstlisting}
#include <iostream>
using namespace std;
\end{lstlisting}

The problem with this
is that it may pull in unwanted functions. For instance:

\begin{block}{Why not `using namespace std'?}
  \label{sl:namespace-std-harm}
  \begin{multicols}{2}
    This compiles, but should not:
    \verbatimsnippet{swapname}
    \vfill\columnbreak
    This gives an error:
    \verbatimsnippet{swapusing}
  \end{multicols}
\end{block}

Even if you use \n{using namespace}, you only do this in a source
file, not in a header file. Anyone using the header would have no idea what
functions are suddenly defined.

\begin{slide}{Big namespace no-no}
  \label{sl:h-no-using}
  Do not put \n{using} in a header file that a user may include.
\end{slide}
